Data-driven approaches to identifying knowledge gaps in marine invertebrate biodiversity

### 2017-03-16 manuscript.Rmd
edit manuscript

With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition. While terrestrial ecosystems have received considerable attention, marine ecosystems are much less understood, and what constitutes the baseline of an healthy and resilient system is not fully characterized. Even basic information such as the number of species that live in the sea is not agreed on, and a significant portion of marine biodiversity remains to be discovered and described. There is therefore a pressing need to improve our knowledge of this environment before species go extinct and its functionning is irreversibly altered. We also have an heterogenous understanding of marine species distributions which limits our understanding of population sizes and connectivity, in turn making the modelization of marine resources and ecosystem resilience less accurate. Until recently, this knowledge was difficult to synthesize because it was scattered and difficult to access. However, recent advances in digitization and public databases efforts have made available the content of natural history collections, information about species occurrences, and genetic data. Given the urgency with which we need to document biodiversity and the limited means available to do so, we argue that we need to build on our existing knowledge to identify gaps, and we need to be more efficient at collecting information to improve our understanding of the natural world.

The incorporation of molecular data into the study of biodiversity has highlighted the pervasiveness of cryptic species, and contributed to the recognition that species have smaller ranges than their taxonomic identity had implied. Notably, despite early criticisms, the wide adoption of DNA barcoding to study marine diversity has invigorated taxonomy. It was also suggested as a possible solution to the widening gap between the increased need for biodiversity inventories while taxonomic expertise is dwindling. By matching DNA barcodes of unidentified species to a reference public database of accurately identified and vouchered specimens, DNA barcoding can greatly speed up the process of species identification by matching them to already identified species. Sequences with no match make it easier to flag specimens that warrant further taxonomic attention.

@@

Recent developments in high-throughput sequencing technologies are giving new applications to these DNA barcode databases. Metabarcoding has been developed to sequence in parallel the "barcoding gene" across the species included in environmental samples such as plankton tows, fish guts, or artificial structure scrappings. However, to accurately identify the species found in these samples, comprehensive DNA barcode databases are needed.

With decreasing sequencing costs, and the prospect of automating sample collection, these technologies will soon be used for ecosystem and biodiversity monitoring. DNA barcoding provides the possibility to link these newly generated data with historical natural history knowledge about species and their distributions. However, the insights we will gain from these approaches are only going to be as good as the level of completeness and accuracy of the information included in  reference databases.

@@

We extracted `r sum(gom_worms$is_marine, na.rm = TRUE)` marine invertebrates species from the faunistic list of the Gulf of Mexico, and `r sum(kozloff_worms$is_marine, na.rm=TRUE)` species from the Salish Sea. Species are unevenly represented in databases, with 25% to 50% of species listed to occur in these locations, that are not represented in digital records of museum collections. Typically a higher proportion of species from the Pacific Coast than the Gulf of Mexico are not represented. Comparatively, all but 2 of the `r nrow(mammals_species_asm)` mammal species known to occur in the US have digitized museum specimens.

The proportion of species that have DNA barcodes available is surprisingly constant across the better studied phyla. About 20% of species known to occur in the Gulf of Mexico and in the Salish Sea have DNA barcode records.  Because geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records)` records analyzed here have no geographical coordinates information associated with them), and many marine invertebrate species are known to be cryptic complexes that are not as widely distributed as previously thought. Therefore specimens identified under the same species but sampled in distant locations, could represent cryptic species. It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that additional samples are collected, to ensure that the extent of their genetic diversity is captured. This high level of cryptic diversity is reflected in the fact that `r stat_bold$n_spp_muti_bin` species out of `r stat_bold$n_spp` include more than 1 BIN. This number is confounded by some taxonomic identification issues as `r stat_bold$n_shared_bins` BINs are shared across more than one species.

@@

The number of species that occurs in the sea remains uncertain.

>>>>

With increasing impact of human activities on the biosphere, we need efficient and accurate approaches to document changes in species ranges and community composition. While terrestrial ecosystems have received considerable attention, the baseline of healthy and resilient marine ecosystems are not well understood. A major hinderance is that we do not have enough information about which species live where, and there is potentially a large portion of marine biodiversity that remains to be discovered and described. Additionally, marine ecosystem monitoring are lacking to detect, and even less understand, how and why communities change through time. There is therefore a pressing need to improve our knowledge of marine systems before species go extinct, and their functionning are irreversibly altered. Our lack of knowledge about marine invertebrates make the modelisation of trophic networks less accurate, as diverse species assemblages are often reduced as simple functional categories that do no account for their ecological differences. Until very recently, our knowledge of biodiversity was scattered among heterogeneous sources (e.g. natural history collections, publications, expert knowledge), and difficult to access. This heterogeneity and the difficulty to accessing this data have made synthesis and the identification of knowledge gaps difficult. However, with the development of linked, publicly accessible, comprehensive databases of natural history records (e.g. iDigBio), species occurrences (e.g., GBIF, OBIS), genetic data (e.g., GenBank, BOLD), and registers of species names (e.g., WoRMS), it is now easier to quantitavely assess the state of our knowledge for biodiversity data. Here, we argue that given the urgency of the biodiversity crisis, we need to develop a data-driven approach to speed up the documentation of biodiversity and build on our existing knowledge to be more efficient at collecting information to improve our understanding of our natural world. Additionally, we argue that our understanding biodiversity in the marine environment and the impact of human activities on communities requires not only to get a more accurate list of the species that inhabits the oceans, but also to understand better their distributions. This can only be done with additional sampling and a commitment to training more taxonomists and experts that can identify the samples collected. To some extant, we are at a time when it has never been easier and cost effective to survey comprehensively the extant of marine biodiversity.

Biodiversity discovery and documentation has been transformed in the 21st century by using molecular data to facilitate species identification, delineate species limits, and more recently to quantify diversity and uncover large scale patterns of biodiversity organization. The wide adoption of DNA barcoding (where homologous DNA sequences are sequenced across the main domains of the Eukaryotes) has unraveled the pervasiveness of cryptic species, invigorating taxonomy and the description of new species since the early 2000s. Unraveling these cryptic species has also contributed to the recognition that species have smaller ranges than their taxonomic identity had implied making them potentially more susceptible to extinction or overexploitation. And despite early criticisms, DNA barcoding has also been suggested as a possible solution to the widening gap between the increased need to document biodiversity while taxonomic expertise is dwindling. Thus, large databases of DNA barcodes have been collaboratively assembled and made pubicly available. By matching DNA barcodes of unidentified species to a reference public database of accurately identified and vouchered specimens, DNA barcoding can greatly speed up species identification. Sequences with no match make it easier to flag specimens that warrant further taxonomic attention.

Barcoding/metabarcoding certainly not perfect methods, but much better than current understanding.

@@

Recent developments in high-throughput sequencing technologies are giving new applications to these DNA barcode databases. Metabarcoding has been developed to sequence in parallel the barcoding gene across the species included in environmental samples such as plankton tows, fish guts, or artificial structure scrappings. With decreasing sequencing costs, and the prospect of automating sample collection, these technologies will soon be used for ecosystem and biodiversity monitoring. DNA barcoding provides the possibility to link these newly generated data with historical natural history knowledge about species and their distributions. However, the insights we will gain from these approaches are only going to be as good as the level of completeness and accuracy of the information included in reference databases. However, to accurately identify the species found in these samples, comprehensive DNA barcode databases are needed. Here, we focus mostly on museum specimens because we can double check their identification and we can potentially extract DNA from them to complete existing databases of DNA barcodes.

@@

We extracted `r sum(gom_worms$is_marine, na.rm = TRUE)` marine invertebrates species from the faunistic list of the Gulf of Mexico, and `r sum(kozloff_worms$is_marine, na.rm=TRUE)` species from the Salish Sea. Species are unevenly represented in databases, with 25% to 50% of species listed to occur in these locations, that are not represented in digital records of museum collections. Typically a higher proportion of species from the Pacific Coast than the Gulf of Mexico are not represented. Comparatively, all but 2 of the `r nrow(mammals_species_asm)` mammal species and `r sum(amphibian_summary$count == 0)` of the `r nrow(amphibian_summary)` amphibian species known to occur in the US have digitized museum specimens.

The proportion of species that have DNA barcodes available is surprisingly constant across the better studied phyla. About 20% of species known to occur in the Gulf of Mexico and in the Salish Sea have DNA barcode records.  Because geographic information is missing for most records (`r format_output(100*stat_bold$p_no_coords)`% of the `r format(stat_bold$n_total_records)` records analyzed here have no geographical coordinates information associated with them), and many marine invertebrate species are known to be cryptic complexes that are not as widely distributed as previously thought. Therefore specimens identified under the same species but sampled in distant locations, could represent cryptic species. It is thus crucial to ensure that accurate geographic information is provided when DNA barcodes are published, and that additional samples are collected, to capture more accurately their geographic range and the extent of their genetic diversity. This high level of cryptic diversity is reflected in the fact that `r stat_bold$n_spp_muti_bin` species out of `r stat_bold$n_spp` include more than 1 BIN. This number is confounded by some taxonomic identification issues as `r stat_bold$n_shared_bins` BINs are shared across more than one species.

@@
